---
title: "Censored regression for multiplex cytokine data"
author: "immunoPlex package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Censored regression for multiplex cytokine data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 5
)
```

## What this vignette covers

The censoring-aware regression tools in immunoPlex:

- the model families and what each one estimates,
- the input scale the fitters expect,
- upper LODs, replicate and plate random effects,
- comparing LOD-handling strategies with `compare_lod_models()`,
- variance-homogeneity testing,
- diagnostics.

For the basic preprocess → fit → diagnose flow, see
`vignette("getting-started")`.

## Available families

| Family | Implementation | Link | Estimand | Censoring handling |
|---|---|---|---|---|
| `gaussian`     | `glmmTMB` | identity | mean difference (log-scale) | substitute upstream (e.g. LOD/2) |
| `gamma`        | `glmmTMB` | log      | ratio of means              | impute at LOD              |
| `tobit`        | `survreg` | identity | ratio of means              | interval `[-∞, LOD]`       |
| `aft`          | `survreg` | log      | ratio of medians            | interval `[ε, LOD]`        |
| `tobit_censreg`| `censReg` | identity | ratio of means              | single LOD only            |
| `auto`         | heuristic | varies   | varies                      | data-driven selection      |

Every fitted model carries an `estimand` field, so you do not have to
remember which family targets means and which targets medians.

## The input scale

All the fitters share one convention:

- **`value`** is on the **log** scale (log concentration),
- **`lod` / `ulod`** stay on the **raw** concentration scale.

`fit_one()` relies on this. It builds Tobit censoring bounds as
`log(LOD)` to match `value`, and the AFT family exponentiates `value`
to return to raw concentrations. Data from `simulate_immunoassay()`
and the `simulate_*` helpers already follows the convention
(`value` is `log(value_raw)`).

`prepare_cytokine_data()` returns `value` on the raw scale, so
log-transform it yourself:

```{r}
library(immunoPlex)

dat1 <- prepare_cytokine_data("IL-6")
dat1$value <- log(dat1$value)
```

## Fitting a single family

```{r}
fit_tobit <- fit_one(
  dat1,
  family = "tobit",
  fixed  = "timepoint*disease + age",
  random = "",         # tobit / aft do not use random effects
  ulod   = TRUE        # use upper-LOD information when present
)

summary(fit_tobit)
```

```{r tobit-censor-aware-plot, fig.cap = "Censor-aware residuals for the Tobit fit. Censored observations are plotted at their interval edges."}
plot(fit_tobit, plot_type = "censor_aware", save_pdf = FALSE)
```

The `auto` selector picks a family from the censoring rate and
skewness:

```{r}
fit <- fit_one(dat1, family = "auto")
fit$family
fit$estimand
```

Its thresholds are options:

```{r eval = FALSE}
options(
  immunoplex.high_censoring_pct = 70,
  immunoplex.high_skewness      = 2
)
```

## Replicate and plate random effects

Multiplex data often has duplicate wells per sample and runs across
several plates. `rep_col` names replicates within a subject, and
`plate_col` names a grouping that crosses subjects.

`prepare_cytokine_data()` carries every metadata column through, so
`replicate` is already available. We add a plate label for
illustration:

```{r}
set.seed(1)
dat1$plate <- sample(paste0("P", 1:4), nrow(dat1), replace = TRUE)

fit_re <- fit_one(
  dat1,
  family    = "gaussian",     # tobit / aft ignore random effects
  fixed     = "timepoint + disease + age",
  random    = "(1|subject_id)",
  rep_col   = "replicate",    # nested:  (1 | subject_id:replicate)
  plate_col = "plate"         # crossed: (1 | plate)
)
fit_re$family
fit_re$converged
```

`rep_col` is nested because the same replicate label on two different
subjects does not mean the same physical replicate. `plate_col` is
flat because plates cross subjects.

Only the families that use `random` (`gamma`, `gaussian`) read these
arguments; `tobit`, `aft` and `tobit_censreg` ignore them. For those,
aggregate replicates first — see `vignette("replicate-qc")`.

Random effects are dropped automatically when the data is too small to
support them, which is configurable:

```{r eval = FALSE}
options(
  fit_one.min_subjects = 30L,   # need >= 30 clusters to keep RE
  fit_one.min_reps     = 3L     # and >= 3 rows per cluster
)
```

## Comparing families

```{r}
models <- fit_models(
  dat1,
  families = c("gaussian", "tobit"),
  ulod     = TRUE,
  random   = ""
)

models$comparison[, c("family", "estimand", "aic", "bic", "delta_aic",
                      "n_cens_lod", "n_cens_ulod")]
```

```{r model-comparison-plot, fig.cap = "AIC ranking for the fitted families."}
plot(models, plot_type = "comparison", save_pdf = FALSE)
```

```{r best-residuals-plot, fig.cap = "Residual diagnostics for the best-AIC family in the comparison."}
plot(models, plot_type = "best_residuals", save_pdf = FALSE)
```

AIC compares models fitted on the same response scale. `gaussian` and
`tobit` both use `value`, so they rank directly. `aft` is fitted on raw
concentrations, so its AIC is offset from the others by a constant and
is not comparable with them.

## Comparing LOD-handling strategies

`compare_lod_models()` puts substitution-based preprocessing against
censoring-aware models on the same data. This function takes `value`
on the **raw** scale, because the substitution rules it tests are
defined in raw units; it log-transforms internally where a family
needs it.

```{r}
dat_raw <- prepare_cytokine_data("IL-6")   # raw scale, not logged

cmp <- compare_lod_models(
  dat_raw,
  families              = c("gamma", "tobit"),
  lod_methods           = c("half", "halfmin"),
  include_log_transform = TRUE
)
cmp$comparison[, c("model", "family", "lod_method", "aic", "delta_aic")]
```

```{r lod-comparison-plot, fig.cap = "Substitution-based preprocessing versus censoring-aware models for the same cytokine."}
plot(cmp, plot_best = TRUE)
```

## Variance-homogeneity testing

For Gaussian fits you can model heterogeneous variance through
`dispformula` and test it against a homogeneous baseline with a
likelihood ratio test:

```{r}
cmp_var <- compare_variance_models(
  data               = dat1,
  fixed              = "timepoint + disease + age",
  random             = "(1|subject_id)",
  dispformula_hetero = ~timepoint
)

cmp_var$preferred_model
cmp_var$comparison[, c("aic_homo", "aic_hetero", "lrt_statistic", "lrt_pvalue")]
```

## Diagnostics

```{r}
# DHARMa scaled residuals, for the glmmTMB families
dharma_diagnostics(fit_re, n_sim = 200)$tests

check_convergence(fit_re)$converged
check_random_effects(fit_re)$icc
```

```{r}
# Influential subjects, via leave-one-out
sens <- leave_one_out_sensitivity(
  data                = dat1,
  fixed               = "timepoint + disease + age",
  target_terms        = "timepointT2",
  random              = "(1|subject_id)",
  influence_threshold = 20,
  progress            = FALSE
)
head(sens[order(-abs(sens$pct_change)), ], 3)
```

```{r}
# Predictive performance for the best model in a set
best_fit <- models$models[[models$best_family]]
kfold_cv_metrics(best_fit, k = 5L)
simulate_pi_coverage(best_fit, nsim = 100L, level = 0.95)
```

## See also

- `vignette("getting-started")` — the workflow end to end
- `vignette("rank-ancova")` — rank-based panel-wide screening
- `vignette("replicate-qc")` — replicate handling and aggregation
- `?fit_one`, `?fit_models`, `?compare_lod_models`,
  `?compare_variance_models`
