---
title: "Getting started with immunoPlex"
author: "immunoPlex package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with immunoPlex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 5
)
```

## What this vignette covers

A short end-to-end tour of the modeling workflow:

1. shape raw multiplex data into the long format used downstream,
2. fit a censoring-aware model for one cytokine,
3. compare model families,
4. run diagnostics.

Other vignettes cover the topic-specific tools:

- `vignette("censored-regression")` — the model families in detail
- `vignette("rank-ancova")` — panel-wide screening with FDR
- `vignette("mcnemar-detection-analysis")` — paired detection frequency
- `vignette("plsda-analysis")` — PLS-DA / OPLS-DA
- `vignette("replicate-qc")` — replicate QC and aggregation

## Setup

```{r}
library(immunoPlex)

data("immunoplex_example", package = "immunoPlex")
str(immunoplex_example, max.level = 1)
```

The bundled example is a 20-analyte panel on 120 subjects, measured at
two timepoints in duplicate wells. `list_cytokines()` summarises the
panel:

```{r}
head(list_cytokines(), 5)
```

## 1. Preprocess

`immuno_preprocess()` joins the expression matrix to the metadata,
attaches per-analyte LOD / ULOD thresholds, and returns long data with
one row per (sample, cytokine).

```{r}
dat <- immuno_preprocess(
  expr        = immunoplex_example$expression,
  meta        = immunoplex_example$metadata,
  lod_lookup  = immunoplex_example$lod_lookup,
  lod_methods = "half"
)

head(dat)
```

`lod_methods` sets what replaces a below-LOD reading in the
`concentration` column — `"half"` (the default) uses LOD/2, and
`"none"` keeps the raw value. The separate `censored` column always
records which rows were below the LOD.

## 2. Fit one model

`prepare_cytokine_data()` pulls a single analyte out of the example
data and builds the columns `fit_one()` expects — `value`, `cens_lod`
and `cens_ulod`:

```{r}
dat1 <- prepare_cytokine_data("IL-6")
head(dat1)
```

`fit_one()` expects `value` on the log scale, so transform it first
(`lod` and `ulod` stay on the raw scale):

```{r}
dat1$value <- log(dat1$value)
```

Now fit. `family = "auto"` picks a family from the censoring rate and
skewness of the data:

```{r}
fit <- fit_one(
  dat1,
  family = "auto",
  fixed  = "timepoint*disease + age",
  random = "(1|subject_id)"
)

fit$family      # the family that was chosen
fit$estimand    # what the coefficients mean
summary(fit)
```

```{r censor-aware-plot, fig.cap = "Censor-aware diagnostic plot. Censored observations are highlighted to show how they enter the likelihood."}
plot(fit, plot_type = "censor_aware", save_pdf = FALSE)
```

## 3. Compare model families

`fit_models()` runs `fit_one()` over several families and returns a
comparison table:

```{r}
models <- fit_models(
  dat1,
  families = c("gaussian", "tobit"),
  ulod     = TRUE,
  random   = ""
)

models$comparison[, c("family", "estimand", "aic", "delta_aic")]
```

```{r model-comparison-plot, fig.cap = "AIC ranking across the fitted families."}
plot(models, plot_type = "comparison", save_pdf = FALSE)
```

```{r best-residuals-plot, fig.cap = "Residual diagnostics for the best-AIC model in the set."}
plot(models, plot_type = "best_residuals", save_pdf = FALSE)
```

Both families here are fit on `value`, so their AIC values sit on the
same scale. `aft` is also available; it is fit on raw concentrations,
so compare it with the others by estimand rather than by AIC.

## 4. Diagnostics

DHARMa residuals and the convergence checks work on the `glmmTMB`
families (`gaussian`, `gamma`):

```{r}
fit_g <- fit_one(
  dat1,
  family = "gaussian",
  fixed  = "timepoint + disease + age",
  random = "(1|subject_id)"
)

diag <- dharma_diagnostics(fit_g, n_sim = 200)
diag$passed_all

check_convergence(fit_g)$converged
check_random_effects(fit_g)$icc
```

Influence and predictive checks work on any `immuno_fit`:

```{r}
sens <- leave_one_out_sensitivity(
  data         = dat1,
  fixed        = "timepoint + disease + age",
  target_terms = "timepointT2",
  random       = "(1|subject_id)",
  progress     = FALSE
)
head(sens[order(-abs(sens$pct_change)), ], 3)

kfold_cv_metrics(fit, k = 5L)
```

## Where to go next

- One cytokine, censoring in depth → `vignette("censored-regression")`
- Whole panel with FDR → `vignette("rank-ancova")`
- Detection frequency between timepoints →
  `vignette("mcnemar-detection-analysis")`
- Multivariate signatures → `vignette("plsda-analysis")`
- Duplicate wells and plates → `vignette("replicate-qc")`
