---
title: "Paired detection-frequency analysis with McNemar's test"
author: "immunoPlex package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Paired detection-frequency analysis with McNemar's test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 5
)
```

## What this vignette covers

`mcnemar_detection()` answers a paired question about *detection*
rather than magnitude: did the proportion of samples above the LOD
change between two timepoints? We cover:

- the basic call and its output,
- the test variants,
- replicate aggregation,
- alluvial plots of detection flow.

## Setup

`simulate_mcnemar_data()` generates a paired dataset:

```{r}
library(immunoPlex)

mc <- simulate_mcnemar_data(
  n_subjects         = 80,
  n_analytes         = 10,
  baseline_detection = 0.4,    # 40% detected at T1
  discordant_rate    = 0.20,   # 20% of subjects change status
  seed               = 1
)

dat <- mc$data
head(dat[, c("subject_id", "timepoint", "cytokine", "cens_lod")])
```

The simulated data carries `cens_lod` (TRUE = below LOD). Data from
`immuno_preprocess()` has the same information in a column named
`censored`, so point `censoring_col` at whichever name your data uses.

## The basic call

```{r}
res <- mcnemar_detection(
  data          = dat,
  subject_col   = "subject_id",
  cytokine_col  = "cytokine",
  timepoint_col = "timepoint",
  censoring_col = "cens_lod",      # detected = !censored
  baseline      = "T1",
  comparison    = "T2",
  method        = "exact",         # or "midp", "noCC"
  fdr_method    = "BH",
  quiet         = TRUE
)

print(res)
```

```{r}
head(res[, c("cytokine", "n_pairs",
             "delta_detection", "delta_ci_lo", "delta_ci_hi",
             "mpor", "p_mcnemar", "q_mcnemar")])
```

### What the output holds

For each cytokine:

- **`delta_detection`** — the percentage-point change in detection,
  with an exact interval in `delta_ci_lo` / `delta_ci_hi`,
- **`mpor`** — matched-pair odds ratio, with its own interval,
- **`rate_upward`** / **`rate_downward`** — transition rates with
  exact intervals,
- **`p_mcnemar`** — p-value from the chosen variant,
- **`q_mcnemar`** — FDR-adjusted across analytes,
- **`cohen_kappa`**, **`detection_pattern`** — supplementary.

## Test variants

```{r eval = FALSE}
# Exact McNemar (default, via exact2x2)
mcnemar_detection(dat, censoring_col = "cens_lod",
                  baseline = "T1", comparison = "T2",
                  method = "exact")

# Mid-p
mcnemar_detection(dat, censoring_col = "cens_lod",
                  baseline = "T1", comparison = "T2",
                  method = "midp")

# No continuity correction, asymptotic Wald interval
mcnemar_detection(dat, censoring_col = "cens_lod",
                  baseline = "T1", comparison = "T2",
                  method = "noCC")
```

## Replicate aggregation

When the data has technical replicates per (subject, timepoint),
`mcnemar_detection()` collapses them to one boolean per pair using
`replicate_agg`:

```{r eval = FALSE}
mcnemar_detection(
  data          = replicated_data,
  censoring_col = "cens_lod",
  baseline      = "T1",
  comparison    = "T2",
  replicate_agg = "majority_vote"   # or "any_detected", "all_detected"
)
```

`aggregate_replicates()` exposes the same rules if you would rather
collapse upstream — see `vignette("replicate-qc")`.

## Custom column names

```{r eval = FALSE}
res <- mcnemar_detection(
  data          = my_data,
  subject_col   = "patient_id",
  cytokine_col  = "marker",
  timepoint_col = "visit",
  baseline      = "Visit1",
  comparison    = "Visit2",
  conf_level    = 0.99
)
```

## Alluvial plots

An alluvial diagram shows how subjects move between detected and
undetected across the two timepoints:

```{r alluvial-single, fig.cap = "Alluvial plot for the most significant cytokine. Each ribbon is a group of subjects following one detection trajectory; ribbon width is proportional to how many."}
plot_mcnemar_alluvial(res)
```

```{r alluvial-grid, fig.cap = "Faceted alluvial plot for the top four cytokines.", fig.height = 6}
plot_mcnemar_alluvial_grid(
  res,
  n_cytokines = 4,
  sort_by     = "significance",   # or "delta_abs", "delta"
  ncol        = 2
)
```

The left bar is detection status at baseline and the right bar at
comparison; flow colour shows the direction of change, and the title
carries the cytokine name and its delta. Both functions take
`baseline_label` and `comparison_label` for the axis names.

## Building a results table

```{r eval = FALSE}
library(dplyr)

table <- res |>
  filter(q_mcnemar < 0.05) |>
  mutate(
    delta_fmt = sprintf("%+.1f [%.1f, %.1f]",
                        delta_detection, delta_ci_lo, delta_ci_hi),
    mpor_fmt  = sprintf("%.2f [%.2f, %.2f]",
                        mpor, mpor_ci_lo, mpor_ci_hi)
  ) |>
  select(cytokine, n_pairs, delta_fmt, mpor_fmt, q_mcnemar)
```

## See also

- `vignette("getting-started")` — the workflow end to end
- `vignette("censored-regression")` — quantitative paired analysis
- `vignette("replicate-qc")` — replicate aggregation rules
- `?mcnemar_detection`, `?plot_mcnemar_alluvial`,
  `?plot_mcnemar_alluvial_grid`

## References

- Fay, M. P., & Lumbard, K. (2021). Confidence intervals for difference
  in proportions for matched pairs compatible with exact McNemar's or
  sign tests. *Statistics in Medicine*, 40(5), 1147–1159.
