---
title: "PLS-DA / OPLS-DA pattern discovery"
author: "immunoPlex package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PLS-DA / OPLS-DA pattern discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 5
)
ropls_available <- requireNamespace("ropls", quietly = TRUE)
knitr::opts_chunk$set(eval = ropls_available)
```

## What this vignette covers

Supervised multivariate pattern discovery: PLS-DA and its orthogonal
variant OPLS-DA, with permutation testing and cross-validation. We
cover:

- preprocessing a wide matrix,
- fitting and validating a model,
- choosing a cross-validation scheme,
- the four plot helpers,
- switching to OPLS-DA.

## Setup

PLS-DA wraps the
[`ropls`](https://bioconductor.org/packages/ropls/) Bioconductor
package. Install it once:

```{r eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("ropls")
```

```{r}
library(immunoPlex)

data("immunoplex_example", package = "immunoPlex")
```

## 1. Preprocess

`plsda_preprocess()` takes a **wide** data frame — one row per sample,
one column per cytokine — plus a per-analyte LOD lookup. It applies
LOD substitution, optional log transformation and optional scaling,
and returns a `plsda_preprocessed` object.

This is a separate entry point from `immuno_preprocess()`, not a step
after it: it wants the raw wide matrix and applies its own LOD rule.
Here we build one from the example data at a single timepoint:

```{r}
expr <- immunoplex_example$expression
meta <- immunoplex_example$metadata
cytokines <- names(expr)

wide <- cbind(sample_id = rownames(expr),
              as.data.frame(expr, check.names = FALSE))
wide <- merge(wide, meta, by = "sample_id")
wide <- subset(wide, timepoint == "T1")

c(rows = nrow(wide), subjects = length(unique(wide$subject_id)))
```

Each subject appears twice, once per replicate well, and PLS-DA needs
one row per sample. `replicate_col` names the column that **groups**
replicates together — the biological sample ID:

```{r}
preprocessed <- plsda_preprocess(
  data          = wide,
  cytokine_cols = cytokines,
  metadata_cols = c("subject_id", "disease", "age"),
  lod_lookup    = immunoplex_example$lod_lookup,
  lod_method    = "half",        # also "halfmin", "sqrt", "lod", "zero"
  log_transform = TRUE,
  scale_data    = TRUE,
  replicate_col = "subject_id",  # average the wells per subject
  verbose       = FALSE
)

dim(preprocessed$expression)
```

`plsda_fit()` has a `rep_col` argument that collapses replicates too,
but it names the opposite thing — the column that *distinguishes*
replicates (`replicate`) — and it runs after scaling rather than
before.

For one-step transforms, `transform` overrides `log_transform` and
`scale_data`:

```{r eval = FALSE}
preprocessed <- plsda_preprocess(
  data          = wide,
  cytokine_cols = cytokines,
  metadata_cols = c("subject_id", "disease"),
  lod_lookup    = immunoplex_example$lod_lookup,
  replicate_col = "subject_id",
  transform     = "log2p1_zscore"   # or "cuberoot_zscore", "pareto"
)
```

## 2. Fit a model

```{r}
model <- plsda_fit(
  preprocessed_data = preprocessed,
  response_var      = "disease",
  n_components      = 2,
  method            = "PLS-DA",   # or "OPLS-DA"
  permutations      = 200,        # 0 to skip permutation testing
  cross_validation  = NULL,       # NULL -> auto, see below
  verbose           = FALSE
)

model$model_stats[, c("R2X(cum)", "R2Y(cum)", "Q2(cum)",
                      "perm_pval_R2Y", "perm_pval_Q2")]
```

The fitted object carries `R2X` and `R2Y` (variance explained), `Q2`
(cross-validated predictive performance), the permutation p-values,
and per-cytokine VIP scores in `model$vip_scores`.

## 3. Cross-validation

`recommend_cv_method()` reports the scheme that `plsda_fit()` picks
for a given sample size:

```{r}
recommend_cv_method(n_total = 75, n_min_group = 20, verbose = TRUE)
```

The defaults, all adjustable through the `*_threshold` and `*_k`
arguments to both functions:

| Sample size | Method |
|---|---|
| `n_total < 50` **or** `n_min_group < 15`     | LOOCV   |
| `50 <= n_total < 100` and all groups >= 15   | 7-fold  |
| `n_total >= 100` and all groups >= 15        | 10-fold |

Override it directly when you want to:

```{r eval = FALSE}
plsda_fit(preprocessed, response_var = "disease",
          cross_validation = "loocv")

plsda_fit(preprocessed, response_var = "disease",
          cross_validation = 7L)

plsda_fit(preprocessed, response_var = "disease",
          loocv_n_threshold     = 75,
          loocv_group_threshold = 20)
```

## 4. Visualise

Four helpers, each returning a ggplot object:

```{r pls-scores-plot, fig.cap = "Score plot: each point is a sample, projected onto the first two latent components."}
plot_pls_scores(model, color_by = "disease", add_ellipse = TRUE)
```

```{r pls-vip-plot, fig.cap = "VIP scores. The dashed line marks VIP = 1."}
plot_pls_vip(model, top_n = 15)
```

```{r pls-loadings-plot, fig.cap = "Loadings: each cytokine's contribution to each latent component."}
plot_pls_loadings(model)
```

```{r pls-biplot, fig.cap = "Biplot: scores and loadings on the same axes."}
plot_pls_biplot(model, color_by = "disease")
```

There is also a `plot()` method on the model, with `plot_type` set to
`"scores"`, `"loadings"`, `"vip"` or `"biplot"`.

## 5. Summary and prediction

```{r}
summary(model)
```

`predict()` projects new samples onto the fitted model:

```{r eval = FALSE}
predict(model, newdata = new_preprocessed)
```

## OPLS-DA

OPLS-DA splits variance into a predictive component and orthogonal
components. Use `method = "OPLS-DA"` for two-class problems; for more
than two classes, use PLS-DA.

```{r eval = FALSE}
opls <- plsda_fit(
  preprocessed_data = preprocessed,
  response_var      = "disease",
  method            = "OPLS-DA",
  n_components      = 2,          # 1 predictive + 1 orthogonal
  permutations      = 200
)

plot_pls_scores(opls, color_by = "disease")
plot_pls_vip(opls, top_n = 15)
```

## See also

- `vignette("getting-started")` — the workflow end to end
- `vignette("replicate-qc")` — aggregating replicates
- `?plsda_preprocess`, `?plsda_fit`, `?recommend_cv_method`
- `?plot_pls_scores`, `?plot_pls_loadings`, `?plot_pls_vip`,
  `?plot_pls_biplot`

## References

- Wold, S., Sjöström, M., & Eriksson, L. (2001). PLS-regression: a
  basic tool of chemometrics. *Chemometrics and Intelligent Laboratory
  Systems*, 58(2), 109–130.
- Trygg, J., & Wold, S. (2002). Orthogonal projections to latent
  structures (O-PLS). *Journal of Chemometrics*, 16(3), 119–128.
- Thévenot, E. A., et al. (2015). Analysis of the human adult urinary
  metabolome variations with age, body mass index, and gender by
  implementing a comprehensive workflow for univariate and OPLS
  statistical analyses. *Journal of Proteome Research*, 14(8),
  3322–3335.
