---
title: "Rank-based ANCOVA for panel-wide cytokine screening"
author: "immunoPlex package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rank-based ANCOVA for panel-wide cytokine screening}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 5
)
```

## What this vignette covers

The rank-based ANCOVA module screens a whole panel with one recipe:
a group effect adjusted for a covariate, an effect size per analyte,
and FDR correction across analytes. We cover:

- single-analyte fitting with `ancova_one()`,
- panel-wide iteration with `ancova_fit()`,
- replicate and plate random intercepts,
- conditional covariate selection,
- tuning the defaults.

## Prepare the data

```{r}
library(immunoPlex)

data("immunoplex_example", package = "immunoPlex")

dat_raw <- immuno_preprocess(
  expr       = immunoplex_example$expression,
  meta       = immunoplex_example$metadata,
  lod_lookup = immunoplex_example$lod_lookup
)
```

`ancova_one()` and `ancova_fit()` fit a linear model that treats each
row as one independent observation. The example data has two
timepoints and two replicate wells per subject, so we reduce it to one
row per subject and analyte first. Here `disease` and `age` are fixed
subject characteristics, so we take a single timepoint and average the
wells:

```{r}
dat_t1 <- subset(dat_raw, timepoint == "T1")

agg <- aggregate_replicates(
  dat_t1,
  rep_col      = "replicate",
  subject_col  = "subject_id",
  cytokine_col = "cytokine",
  value_col    = "concentration",
  rule         = "mean"
)

# aggregate_replicates() returns the grouping keys and the value only,
# so join the subject-level metadata back on
subject_meta <- unique(dat_t1[, c("subject_id", "disease", "age", "preexist")])
dat_long     <- merge(agg, subject_meta, by = "subject_id")

c(rows = nrow(dat_long), subjects = length(unique(dat_long$subject_id)))
```

To keep the replicate rows instead, pass `rep_col` — see "Replicate
and plate structure" below.

## Single analyte: `ancova_one()`

```{r}
dat_one <- subset(dat_long, cytokine == "IL-6")

fit <- ancova_one(
  data            = dat_one,
  outcome         = "concentration",
  group           = "disease",
  covariate       = "age",       # rank-transformed internally
  outlier_removal = TRUE,        # IQR fence on outcome + covariate
  iqr_multiplier  = 3,
  conf_level      = 0.95,
  log_validate    = TRUE         # also fit a log-scale validation model
)

print(fit)
```

```{r ancova-one-plot, fig.cap = "Rank-based ANCOVA diagnostics for one cytokine: residuals-versus-fitted and quantile-quantile plots."}
plot(fit)
```

What the engine does, in order:

1. apply IQR outlier removal on `outcome` and `covariate`,
2. drop binary `conditional_covariates` below `prevalence_threshold`,
3. rank-transform `outcome` and `covariate`,
4. fit `lm(rank(outcome) ~ rank(covariate) + group + covariates)`,
5. compute partial omega-squared effect sizes per term,
6. optionally fit a log-scale model and report concordance with the
   rank-based fit.

## Panel-wide: `ancova_fit()`

```{r}
result <- ancova_fit(
  data          = dat_long,
  outcome       = "concentration",
  group         = "disease",
  covariate     = "age",
  analyte_col   = "cytokine",
  fdr_method    = "BH",
  fdr_threshold = 0.05,
  quiet         = TRUE
)

result$n_analytes        # total fitted
result$n_significant     # below fdr_threshold
head(result$results[, c("analyte", "n_obs", "omega_sq_partial",
                        "group_p", "q_value")], 5)
```

```{r ancova-set-plot, fig.cap = "Panel-wide summary: per-analyte effect sizes (partial omega-squared) with significance flags."}
plot(result)
```

The result is an `immuno_ancova_set`. The per-analyte `immuno_ancova`
objects are kept under `$models`, so you can drill into any cytokine:

```{r ancova-drilldown-plot, fig.cap = "Diagnostics for a single cytokine pulled from the panel-wide result."}
plot(result$models[["IFN-g"]])
```

## Replicate and plate structure

Pass `rep_col`, `plate_col`, or both, using the un-aggregated data:

```{r eval = FALSE}
set.seed(1)
dat_t1$plate <- sample(paste0("P", 1:4), nrow(dat_t1), replace = TRUE)

result_re <- ancova_fit(
  data      = dat_t1,          # replicate-level rows
  outcome   = "concentration",
  group     = "disease",
  covariate = "age",
  rep_col   = "replicate",     # nested: (1 | subject_id:replicate)
  plate_col = "plate"          # crossed: (1 | plate)
)
```

Both columns must already exist in the data — `ancova_fit()` stops
with an error if either name is missing. The example data has no plate
layout, so the plate column above is synthesized for illustration.

With `rep_col` or `plate_col` set, each per-analyte fit switches from
`lm()` to `glmmTMB::glmmTMB()` (Gaussian) with the matching random
structure. For k-level groups the omnibus test becomes a likelihood
ratio test rather than the Type-I F, so `omnibus_f` is `NA` and
`omnibus_p` carries the LRT p-value.

## Conditional covariate selection

Binary covariates that are rare in the data — a comorbidity present in
one or two subjects, say — can be dropped automatically:

```{r}
covars <- choose_covariates(
  data           = dat_long,
  base_covars    = c("disease", "age"),
  candidates     = c("preexist"),
  rare_threshold = 0.03
)
covars
```

`ancova_one()` and `ancova_fit()` apply this rule internally through
their `conditional_covariates` argument. You can also call
`choose_covariates()` directly when building your own formula.

## Tuning the defaults

The ANCOVA defaults are exposed as options:

```{r eval = FALSE}
options(
  immunoplex.ancova_iqr_multiplier       = 3,
  immunoplex.ancova_min_n                = 12L,
  immunoplex.ancova_fdr_method           = "BH",
  immunoplex.ancova_fdr_threshold        = 0.05,
  immunoplex.ancova_prevalence_threshold = 0.03
)
get_immunoplex_config()$ancova
```

## See also

- `vignette("getting-started")` — the workflow end to end
- `vignette("censored-regression")` — parametric single-analyte models
- `vignette("replicate-qc")` — replicate handling
- `?ancova_one`, `?ancova_fit`, `?choose_covariates`
