---
title: "Replicate QC and aggregation"
author: "immunoPlex package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replicate QC and aggregation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 5
)
```

## What this vignette covers

Multiplex assays (Luminex, BioRad, MSD) usually run each biological
sample in duplicate or triplicate wells, sometimes across several
plates. immunoPlex provides two tools for that:

- **`flag_replicate_outliers()`** — flag, or drop, discordant
  replicate pairs using CV and MAD criteria, with optional whole-well
  failure detection,
- **`aggregate_replicates()`** — collapse replicates to one row per
  (subject, cytokine, timepoint).

## Which downstream methods need aggregation

| Downstream method | Replicates |
|---|---|
| `fit_one`, `fit_models` (Gamma / Gaussian) | keep them; pass `rep_col` |
| `fit_one`, `fit_models` (Tobit / AFT)      | aggregate first — these fitters do not use random effects |
| `ancova_one`, `ancova_fit`                 | either — pass `rep_col`, or aggregate first |
| `mcnemar_detection`                        | handled internally through `replicate_agg` |
| `plsda_fit`                                | aggregate — `plsda_preprocess(replicate_col=)` or `plsda_fit(rep_col=)` |

## Setup

```{r}
library(immunoPlex)

data("immunoplex_example", package = "immunoPlex")

dat_long <- immuno_preprocess(
  expr        = immunoplex_example$expression,
  meta        = immunoplex_example$metadata,
  lod_lookup  = immunoplex_example$lod_lookup,
  lod_methods = "none"    # keep raw values so the CV reflects the wells
)
```

`sample_id_col` names the unit whose wells are replicates of one
another. In this dataset that is a subject **at one timepoint**, so we
build a key for it — passing `subject_id` alone would treat T1 and T2
as four replicates of the same sample:

```{r}
dat_long$bio_sample <- paste(dat_long$subject_id, dat_long$timepoint,
                             sep = "_")

table(dat_long$replicate)
```

## Flagging discordant pairs

```{r}
flagged <- flag_replicate_outliers(
  data             = dat_long,
  sample_id_col    = "bio_sample",     # subject x timepoint
  replicate_id_col = "sample_id",      # unique well ID
  cytokine_col     = "cytokine",
  value_col        = "concentration",
  group_cols       = "timepoint",      # population used for the MAD reference
  cv_threshold     = 25,               # typical Luminex QC: 15-25%
  mad_threshold    = 3,
  flag_rule        = "either",         # "cv" | "mad" | "either" | "both"
  action           = "flag",           # annotate only
  verbose          = FALSE
)

mean(flagged$replicate_discordant, na.rm = TRUE)   # fraction flagged
```

```{r cv-distribution-plot, fig.cap = "Within-pair coefficient of variation across all (sample, cytokine) replicate groups. The dashed line marks cv_threshold = 25%."}
library(ggplot2)
ggplot(unique(flagged[, c("bio_sample", "cytokine", "replicate_cv")]),
       aes(x = replicate_cv)) +
  geom_histogram(bins = 60, fill = "steelblue", alpha = 0.85) +
  geom_vline(xintercept = 25, linetype = "dashed", color = "firebrick") +
  coord_cartesian(xlim = c(0, 200)) +
  labs(x = "Within-pair CV (%)", y = "Number of replicate groups") +
  theme_minimal()
```

### The rules

- **CV** (`flag_rule = "cv"`) — flag pairs whose within-pair
  coefficient of variation exceeds `cv_threshold`.
- **MAD** (`"mad"`) — flag pairs whose within-pair absolute difference
  exceeds `mad_threshold ×` the MAD of all within-pair differences for
  that analyte, within `group_cols` if given.
- **`"either"`** flags when either rule fires; **`"both"`** only when
  both do.

### The actions

- `"flag"` — annotate only, drop nothing.
- `"drop_both"` — remove both replicates of a flagged pair.
- `"drop_farther"` — keep the replicate closer to the analyte's
  population median, remove the other.
- `"winsorize"` — pull the farther value toward the closer one.

## Whole-well failure detection

With `well_detection = "sample"` the function also flags entire wells
that fail across many analytes, rather than individual analyte pairs:

```{r eval = FALSE}
flagged <- flag_replicate_outliers(
  data             = dat_long,
  sample_id_col    = "bio_sample",
  replicate_id_col = "sample_id",
  cytokine_col     = "cytokine",
  value_col        = "concentration",
  lod_col          = "lod",
  well_detection   = "sample",
  well_params      = list(
    well_threshold         = 0.3,   # >= 30% of analytes flagged -> well failure
    sign_bias_threshold    = 0.7,   # directional if |mean(sign(z))| above this
    lod_fraction_threshold = 0.5,   # >= 50% at the LOD -> dropout failure
    score_ratio_threshold  = 1.5
  )
)
```

This adds the columns `well_failure`, `well_failure_type`,
`well_score`, `well_sign_bias` and `well_lod_fraction`.

## Aggregating replicates

```{r}
agg <- aggregate_replicates(
  data          = dat_long,
  rep_col       = "replicate",
  subject_col   = "subject_id",
  cytokine_col  = "cytokine",
  timepoint_col = "timepoint",
  value_col     = "concentration",
  rule          = "mean",
  keep_sd       = TRUE
)

head(agg)
```

The result holds the grouping keys and the aggregated value. Metadata
columns such as `disease` or `age` are not carried through, so join
them back from a subject-level table when you need them:

```{r}
subject_meta <- unique(dat_long[, c("subject_id", "disease", "age")])
agg_full     <- merge(agg, subject_meta, by = "subject_id")
names(agg_full)
```

Detection status aggregates with the same rules `mcnemar_detection()`
uses:

```{r}
agg_det <- aggregate_replicates(
  data          = dat_long,
  rep_col       = "replicate",
  subject_col   = "subject_id",
  cytokine_col  = "cytokine",
  timepoint_col = "timepoint",
  censoring_col = "censored",       # TRUE = below LOD
  rule          = "majority_vote"   # or "any_detected", "all_detected"
)

table(agg_det$censored)
```

### The rules

- **Continuous** — `"mean"` (default), `"geomean"` (needs strictly
  positive values), `"median"`.
- **Boolean** — `"majority_vote"` sends 50/50 ties to `FALSE` through
  banker's rounding; `"any_detected"` calls a sample detected if any
  well saw it; `"all_detected"` requires every well to agree.

Edge cases: a single observation per group makes all rules the
identity; fully missing groups return `NA`, including for
`"all_detected"`; and `"geomean"` returns `NA` with a warning if any
replicate is not positive.

## A typical pipeline

```{r eval = FALSE}
# 1. QC the wells
clean <- flag_replicate_outliers(
  dat_long,
  sample_id_col = "bio_sample", replicate_id_col = "sample_id",
  cytokine_col  = "cytokine",   value_col = "concentration",
  flag_rule = "either", action = "drop_farther"
)

# 2a. Into PLS-DA or McNemar: aggregate
agg <- aggregate_replicates(
  clean, rep_col = "replicate",
  subject_col = "subject_id", cytokine_col = "cytokine",
  timepoint_col = "timepoint",
  value_col = "concentration", rule = "geomean"
)

# 2b. Into ancova_fit: keep replicates, pass rep_col
res <- ancova_fit(clean, outcome = "concentration", group = "disease",
                  covariate = "age", analyte_col = "cytokine",
                  rep_col = "replicate")

# 2c. Into fit_one: one cytokine, renamed to the columns it expects
one <- subset(clean, cytokine == "IL-6")
one$value     <- log(one$concentration)
one$cens_lod  <- one$censored
one$cens_ulod <- !is.na(one$ulod) & one$concentration > one$ulod

fit <- fit_one(one, family = "gaussian",
               fixed = "timepoint + disease + age",
               random = "(1|subject_id)",
               rep_col = "replicate")
```

Step 2c is the bridge between the two halves of the package.
`immuno_preprocess()` gives panel-wide long data with a
`concentration` column, while `fit_one()` fits one analyte and expects
`value` on the log scale plus `cens_lod` / `cens_ulod`. See
`vignette("censored-regression")` for that convention.

## See also

- `vignette("getting-started")` — the workflow end to end
- `vignette("censored-regression")` — random effects in parametric fits
- `vignette("rank-ancova")` — random effects in `ancova_*`
- `vignette("mcnemar-detection-analysis")` — boolean replicate rules
- `?flag_replicate_outliers`, `?aggregate_replicates`
